---
title: "Game Probability"
output: html_document
date: "2025-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(purrr)
library(glmnet)
```


# IMPLEMENT BAYESIAN APPROACH:
Use pre-match probability distribution as prior, match results as likelihood to inform posterior

# INSERT ELEMENT OF MOMENTUM:
Treat each set independently, include some factor of momentum on point-level probabilities


WAYS TO IMPROVE THIS PROCESS
1. Find a better way to impute libero on a match-by-match basis
2. Shrink ridge APM values on a position-level basis
3. Have different RAPMs for players in different situations (front row, back row, serving, in receive vs serving)
4. Only include players as a '1' if they touch the ball at some point in the rally

```{r}
# Impute when liberos are on the court
add_liberos <- function(data) {

  player_ids <- data %>%
    group_by(player_id, position) %>%
    summarize(n = n()) %>%
    group_by(player_id) %>%
    slice_max(n, with_ties = FALSE) %>%
    ungroup() %>%
    select(-n) %>%
    rename(pos = position)
  
  shortenings <- data %>%
    select(team, team_short) %>%
    distinct() %>%
    rename(short = team_short)
  
  liberos <- data %>%
    select(team_short) %>%
    distinct() %>%
    filter(!is.na(team_short), !team_short %in% c("southflorida", "wichitastate", "tulsa", "pacific", "westvirginia", "washingtonstate", "memphis", "lmu")) %>%
    rowwise() %>%
    mutate(libero = generate_starting_lineup(power, team_short)$lineup %>% filter(pos == "L") %>% select(player_id) %>% pull()) %>%
    left_join(shortenings, by = c("team_short" = "short")) %>%
    select(team, libero)
  
  data <- data %>%
    mutate(across(
      c(home_player_id1, home_player_id2, home_player_id3, 
        home_player_id4, home_player_id5, home_player_id6,
        visiting_player_id1, visiting_player_id2, visiting_player_id3, 
        visiting_player_id4, visiting_player_id5, visiting_player_id6),
      as.character
    )) %>%
    left_join(player_ids, by =c("home_player_id5" = "player_id")) %>% rename(home_pos5 = pos) %>%
    left_join(player_ids, by =c("home_player_id6" = "player_id")) %>% rename(home_pos6 = pos) %>%
    left_join(player_ids, by =c("home_player_id1" = "player_id")) %>% rename(home_pos1 = pos) %>%
    left_join(player_ids, by =c("visiting_player_id5" = "player_id")) %>% rename(visiting_pos5 = pos) %>%
    left_join(player_ids, by =c("visiting_player_id6" = "player_id")) %>% rename(visiting_pos6 = pos) %>%
    left_join(player_ids, by =c("visiting_player_id1" = "player_id")) %>% rename(visiting_pos1 = pos) %>%
    left_join(liberos, by = c("home_team" = "team")) %>% rename(home_libero = libero) %>%
    left_join(liberos, by = c("visiting_team" = "team")) %>% rename(visiting_libero = libero) %>%
    mutate(
      home_player_id5 = ifelse(home_pos5 == "Middle", home_libero, home_player_id5),
      home_player_id6 = ifelse(home_pos6 == "Middle", home_libero, home_player_id6),
      visiting_player_id5 = ifelse(visiting_pos5 == "Middle", visiting_libero, visiting_player_id5),
      visiting_player_id6 = ifelse(visiting_pos6 == "Middle", visiting_libero, visiting_player_id6)
    ) %>%
    group_by(match_id, point_id) %>%
    mutate(serve_player_id = ifelse(skill[row_number() == 1] == "Serve", player_id[skill == "Serve" & row_number() == 1], NA)) %>%
    ungroup() %>%
    mutate(
      home_player_id1 = case_when(
        home_pos1 != "Middle" ~ home_player_id1,
        home_pos1 == "Middle" & serving_team == visiting_team ~ home_libero,
        home_pos1 == "Middle" & serving_team == home_team ~ serve_player_id,
        TRUE ~ NA
      ),
      visiting_player_id1 = case_when(
        visiting_pos1 != "Middle" ~ visiting_player_id1,
        visiting_pos1 == "Middle" & serving_team == home_team ~ visiting_libero,
        visiting_pos1 == "Middle" & serving_team == visiting_team ~ serve_player_id,
        TRUE ~ NA
      )
    ) %>%
    select(-home_pos5, -home_pos6, -home_pos1, -visiting_pos5, -visiting_pos6, -visiting_pos1, -home_libero, -visiting_libero, -serve_player_id)

  return(data)
}

# Calculate RAPM dataframe
get_rapm <- function(data) {
  
  # try removing games involving a team that has <=3 games in the data
  rare_teams <- data %>%
    group_by(match_id, team) %>%
    summarize(n()) %>%
    group_by(team) %>%
    summarize(count = n()) %>%
    filter(count <= 3)
  
  rare_matches <- data %>%
    filter(team %in% rare_teams$team) %>%
    group_by(match_id) %>%
    summarize(count = n())
  
  # select columns and add point winner (Y in linear regression)
  apm_points <- data %>%
    filter(!match_id %in% rare_matches$match_id) %>%
    select(match_id, point_id,
           home_player_id1, home_player_id2, home_player_id3, home_player_id4, home_player_id5, home_player_id6,
           visiting_player_id1, visiting_player_id2, visiting_player_id3, 
           visiting_player_id4, visiting_player_id5, visiting_player_id6,
           point_won_by, home_team, visiting_team) %>%
    distinct() %>%
    mutate(winner = ifelse(point_won_by == home_team, 1, 0)) %>%
    select(-point_won_by, -home_team, -visiting_team) %>%
    filter(!is.na(winner))
  
  # elongate the df, value of 1 for home players or -1 for visiting players
  apm_long <- 
    apm_points %>%
      pivot_longer(cols = starts_with("home_player_id"), names_to = "role", values_to = "player_id") %>%
      mutate(value = 1) %>% # Home players get +1
    bind_rows(
      apm_points %>%
        pivot_longer(cols = starts_with("visiting_player_id"), names_to = "role", values_to = "player_id") %>%
        mutate(value = -1) # Visiting players get -1
    ) %>%
    select(match_id, point_id, player_id, value, winner) %>%
    filter(player_id != "unknown player") %>%
    mutate(
      plus_minus = case_when(
        value == 1 & winner == 1 ~ 1,
        value == 1 & winner == 0 ~ -1,
        value == -1 & winner == 1 ~ -1,
        value == -1 & winner == 0 ~ 1,
        TRUE ~ NA)
    ) %>%
    mutate(player_id = as.character(player_id))
  
  # find combinations of match, point, player, winner, and plus-minus to get rid of them
  duplicates <- apm_long %>%
    group_by(match_id, point_id, player_id, winner, plus_minus) %>%
    summarize(count = n()) %>%
    filter(count == 2)

  # get rid of duplicates
  apm_long <- apm_long %>% anti_join(duplicates, by = c("match_id", "point_id"))
  
  # widen the df to get one row per point
  apm_wide <- apm_long %>%
    select(-plus_minus) %>%
    pivot_wider(names_from = player_id, values_from = value, values_fill = 0) %>%
    mutate(total = rowSums(select(., -(1:3)), na.rm = TRUE)) %>%
    select(total, everything()) %>%
    select(match_id, point_id, winner, total, everything())

  # build ridge regression
  X <- as.matrix(apm_wide %>% select(-match_id, -point_id, -winner, -total))
  y <- apm_wide$winner
  ridge_apm <- cv.glmnet(X, y, family = "binomial", alpha = 0)
  
  # grab coefficients
  ridge_coefs <- coef(ridge_apm, s = "lambda.1se")
  
  # match players with ids
  player_ids <- data %>%
    group_by(player_id, player_name, team_short, position) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    group_by(player_id, player_name, team_short) %>%
    slice(1)
  
  # find sample sizes
  player_sample <- apm_long %>%
    mutate(player_id = as.character(player_id)) %>%
    group_by(player_id) %>%
    summarize(points = n())
  
  # put it all together
  rapm <- data.frame(
      id = gsub("`","",rownames(ridge_coefs)), 
      ridge_APM = as.numeric(ridge_coefs)
    ) %>%
    filter(id != "(Intercept)") %>%
    filter(!is.na(ridge_APM)) %>%
    left_join(player_ids, by = c("id" = "player_id")) %>%
    left_join(player_sample, by = c("id" = "player_id")) %>%
    select(id, player_name, team_short, position, ridge_APM, points)
  
  return(rapm)

}

# Shrink RAPM values based on sample size and position
shrink_rapm <- function(rapm) {
  
  # Step 1: compute position-specific shrinkage factors (excluding "unsure")
  pos_shrink <- rapm %>%
    filter(position != "unsure") %>%
    group_by(position) %>%
    summarize(
      y_bar = weighted.mean(ridge_APM, points),
      scale_factor = median(points)
    ) %>%
    ungroup()
  
  # Step 2: pull libero’s shrinkage values
  libero_vals <- pos_shrink %>% filter(position == "libero")
  
  # Step 3: join shrinkage params to all players
  rapm <- rapm %>%
    left_join(pos_shrink, by = "position") %>%
    mutate(
      # if unsure, use libero values instead
      y_bar = ifelse(position == "unsure", libero_vals$y_bar, y_bar),
      scale_factor = ifelse(position == "unsure", libero_vals$scale_factor, scale_factor),
      
      shrink_weight = 1 - exp(-points / scale_factor),
      RAPM = shrink_weight * ridge_APM + (1 - shrink_weight) * y_bar
    ) %>%
    select(-y_bar, -scale_factor)
  
  return(rapm)
}

# Combine these
generate_rapm <- function(data) {
  libero_data <- add_liberos(data)
  rapm <- get_rapm(libero_data)
  rapm <- shrink_rapm(rapm)
  
  list(
    libero_data = libero_data,
    rapm = rapm
  )
}

# Evaluate RAPMs
evaluate_rapm <- function(rapm) {
  # plot how quickly this shrinkage occurs
  shrinkage_rate <- ggplot(rapm, aes(x = points, y = shrink_weight)) +
    geom_point() +
    theme_minimal() +
    labs(title = "Shrinkage of Ridge APM", x = "Sample Size", y = "Shrinkage Weight") +
    theme(plot.title = element_text(hjust = 0.5))
  
  # plot the change of RAPM values
  shrinkage_effect <- ggplot(rapm, aes(x = ridge_APM, y = RAPM, color = points)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    geom_point() +
    scale_color_viridis_c() +
    theme_minimal() +
    labs(title = "Shrinkage of Ridge APM", x = "Original Ridge APM", y = "RAPM After Shrinkage") +
    theme(plot.title = element_text(hjust = 0.5))
  
  # evaluate RAPM success using All-American Awards
  npoy_finalists <- rapm %>%
    filter(player_name %in% c("Olivia Babcock", "Brooklyn Deleye", "Sarah Franklin", "Lexi Rodriguez")) %>%
    summarize(mean(RAPM)) %>%
    pull()
  
  first_team <- rapm %>%
    filter(player_name %in% c("Olivia Babcock", "Brooklyn Deleye", "Rachel Fairbanks", "Sarah Franklin", "Andi Jackson", "Kennedy Martin", "Kami Miner", "Jess Mruzik", "Melanie Parra", "Lexi Rodriguez", "Elia Rubin", "Elena Scott", "Torrey Stafford", "Kendra Wait")) %>%
    summarize(mean(RAPM)) %>%
    pull()
  
  second_team <- rapm %>%
    filter(player_name %in% c("Nina Cajic", "Raven Colvin", "Mimi Colyer", "Anna Debeer", "Sami Francis", "Emma Grome", "Eva Hudson", "Bre Kelley", "Logan Lednicky", "Harper Murray", "Bergen Reilly", "Madisen Skinner", "Izzy Starck",
                              "Norah Sis")) %>%
    summarize(mean(RAPM)) %>%
    pull()
  
  third_team <- rapm %>%
    filter(player_name %in% c("Charlie Fuerbringer", "Flormarie Heredia", "Audrey Koenig", "Khori Louis", "Onye Ofoegbu", "Alexis Shelton", "Naya Shime", "Raina Terry", "Mia Tuaniga", "Camryn Turner", "Argentina Ung", "Mychael Vernon",
                              "Lexie Almodovar", "Ava Martin")) %>%
    summarize(mean(RAPM)) %>%
    pull()
  
  honorable_mention <- rapm %>%
    filter(player_name %in% c("Ally Batenhorst", "Bianca Bertolino", "Ifenna Cos-Okpalla", "Cara Cresse", "Celia Cullen", "Geli Cyr", "Kamryn Gibadlo", "Cam Hannah", "Julia Hanson", "Jordan Iliff", "Claire Jeter", "Caroline Kerr", "Cheridyn Leverette", "Claire Little", "Grace Lopez", "Charitie Luper", "Elise McGhee", "Michelle Ohwobete", "Toyosi Onabanjo", "Maya Sands", "Melanie Shaffmaster", "Mary Shroll", "Taylor Trammell", "Nia Washington", "Riley Whitesides", "Kira Fallert",
                              "Caylen Alexander", "Yadhira Anchante", "Maria Clara Andrade", "Milan Bayless", "Maddy Bilinovic", "Kamryn Chaney", "Syd Cole", "Kaylee Cox", "Macy Daufeldt", "Alexa Edwards", "Skye Ekes", "Brianna Ford", "Elena Giacomini", "Elise Goetzinger", "Tori Hester", "Cicily Hidalgo", "Lauryn Hovey", "Tsvetelina Ilieva", "Madolyn Isringhausen", "Malaya Jones", "shelby Kent", "Patrycja Lagida", "Dylan Maberry", "Rya McKinnon", "Kylee Owens", "Panna Ratkai", "Amanda Rice", "Kaya Weaver", "Sylvie Zgnoc")) %>%
    summarize(mean(RAPM)) %>%
    pull()
  
  all_americans <- ggplot(rapm, aes(RAPM)) +
    geom_histogram(fill = "steelblue", color = "black") +
    geom_vline(xintercept = npoy_finalists, color = "black", linetype = "dashed", linewidth = 1.25) +
    geom_vline(xintercept = first_team, color = "gold", linetype = "dashed", linewidth = 1.25) +
    geom_vline(xintercept = second_team, color = "#C0C0C0", linetype = "dashed", linewidth = 1.25) +
    geom_vline(xintercept = third_team, color = "#CD7F32", linetype = "dashed", linewidth = 1.25) +
    geom_vline(xintercept = honorable_mention, color = "firebrick", linetype = "dashed", linewidth = 1.25) +
    labs(title = "Distribution of RAPM Values", x = "RAPM Value", y = "") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  list(
    shrinkage_rate = shrinkage_rate,
    shrinkage_effect = shrinkage_effect,
    all_americans = all_americans
  )
}
```

```{r}
# get a list of players that touched the ball during each rally
touches <- data %>%
  select(match_id, point_id, player_id) %>%
  distinct() %>%
  mutate(touch = 1)

# only keep the players in each point that touched the ball in that rally
apm_long <- apm_long %>%
  left_join(touches, by = c("match_id", "point_id", "player_id")) %>%
  filter(touch == 1) %>%
  select(-touch)
```


```{r}
rapm_result <- generate_rapm(power)
rapm <- rapm_result$rapm
print(head(rapm_result$libero_data))
print(head(rapm_result$rapm))
```

```{r}
rapm_eval <- evaluate_rapm(rapm_result$rapm)
rapm_eval$shrinkage_rate
rapm_eval$shrinkage_effect
rapm_eval$all_americans
```



WAYS TO IMPROVE THIS PROCESS
1. 

```{r}
# Identify which players were on court and their positions for a team
get_on_court_data <- function(data, team_name) {
  data %>%
    mutate(player_id = as.character(player_id)) %>%  # <-- FIX here
    filter(team == team_name) %>%
    select(match_id, point_id, starts_with("home_player_id"), starts_with("visiting_player_id"),
           home_team, visiting_team) %>%
    pivot_longer(cols = starts_with("home_player_id") | starts_with("visiting_player_id"),
                 names_to = "player_slot", values_to = "player_id") %>%
    distinct(match_id, point_id, player_id) %>%
    mutate(player_id = as.character(player_id)) %>%  # <-- ensure type match before join
    left_join(data %>%
                mutate(player_id = as.character(player_id)) %>%
                select(match_id, point_id, player_id, position) %>%
                distinct(), 
              by = c("match_id", "point_id", "player_id")) %>%
    filter(!is.na(position)) %>%
    distinct(match_id, point_id, player_id, position)
}

# Detect system type (5-1 vs 6-2)
detect_system <- function(on_court_data, team_name) {
  # Filter to the target team
  team_on_court <- on_court_data %>%
    filter(team_short == team_name, position == "Setter")

  # Count number of rallies each setter was on court
  setter_usage <- team_on_court %>%
    count(player_id, name = "rallies") %>%
    mutate(share = rallies / sum(rallies))

  if (nrow(setter_usage) == 1 || max(setter_usage$share) > 0.8) {
    return("5-1")
  } else if (
    nrow(setter_usage) == 2 &&
    all(setter_usage$share > 0.3 & setter_usage$share < 0.7)
  ) {
    return("6-2")
  } else {
    return("Unknown")
  }
}

# Build the starting lineup based on the identified system type
build_lineup <- function(data, team_name, system) {
  team_data <- data %>%
    filter(team_short == team_name) %>%
    mutate(player_id = as.character(player_id))

  # Count player usage by position
  top_by_position <- team_data %>%
    count(position, player_id, player_name, sort = TRUE) %>%
    group_by(position) %>%
    arrange(desc(n)) %>%
    ungroup()

  # Define how many players per position for each system
  position_quota <- switch(system,
  "5-1" = c(Setter = 1, Outside = 2, Opposite = 1, Middle = 2, "L/DS" = 3),
  "6-2" = c(Setter = 2, Outside = 2, Opposite = 2, Middle = 2, "L/DS" = 3),
  {
    message("Unknown system — defaulting to 5–1")
    c(Setter = 1, Outside = 2, Opposite = 1, Middle = 2, "L/DS" = 3)
  }
  )

  # Build lineup by taking top N from each position
  lineup <- map_df(names(position_quota), function(pos) {
    n_needed <- position_quota[[pos]]
    
    top_by_position %>%
      filter(position == pos, n > 0.1 * max(n)) %>%
      slice_head(n = n_needed) %>%
      mutate(pos = case_when(
        pos == "Setter" ~ paste0("S", row_number()),
        pos == "Outside" ~ paste0("OH", row_number()),
        pos == "Opposite" ~ paste0("OPP", row_number()),
        pos == "Middle" ~ paste0("MB", row_number()),
        pos == "L/DS" & row_number() == 1 ~ "L",
        pos == "L/DS" & row_number() > 1 ~ paste0("DS", row_number() - 1),
        TRUE ~ pos  # fallback
      )) %>%
      select(position, pos, player_id, player_name, n)
  })
  
  # Only modify if system is 6-2
  if (system == "6-2") {
    servers <- data %>%
      filter(team_short == team_name, skill == "Serve") %>%
      distinct(player_id)

    # Check whether each relevant player is in servers
    opp1_id <- lineup %>% filter(pos == "OPP1") %>% pull(player_id)
    opp2_id <- lineup %>% filter(pos == "OPP2") %>% pull(player_id)
    s1_id   <- lineup %>% filter(pos == "S1") %>% pull(player_id)
    s2_id   <- lineup %>% filter(pos == "S2") %>% pull(player_id)

    s1_serves   <- s1_id %in% servers$player_id
    s2_serves   <- s2_id %in% servers$player_id
    opp1_serves <- opp1_id %in% servers$player_id
    opp2_serves <- opp2_id %in% servers$player_id

    if ((s1_serves && opp1_serves) || (s2_serves && opp2_serves)) {
      lineup <- lineup %>%
        mutate(pos = case_when(
          pos == "OPP1" ~ "OPP2_temp",
          pos == "OPP2" ~ "OPP1",
          pos == "OPP2_temp" ~ "OPP2",
          TRUE ~ pos
        )) %>%
        mutate(pos = ifelse(pos == "OPP2_temp", "OPP2", pos))
    }
  }

  return(lineup)
}

# Find the 6 most common servers for a team
get_servers <- function(data, team_name) {
  data %>%
    filter(team_short == team_name, skill == "Serve") %>%
    count(player_id, player_name, sort = TRUE) %>%
    slice_head(n = 6)
}

# Define which pin hitters play back row as well
get_back_row <- function(data, team_name) {
  players <- power %>%
    filter(team_short == team_name) %>%
    select(player_id, player_name) %>%
    distinct()
    
  power %>%
    mutate(home_team_short = ifelse(team == home_team, team_short, NA)) %>%
    filter(home_team_short == team_name) %>%
    select(match_id, point_id, starts_with("home_player_id")) %>%
    pivot_longer(cols = starts_with("home_player_id"),
                 names_to = "rotation", values_to = "rotation_player_id") %>%
    distinct() %>%
    mutate(rotation_player_id = as.character(rotation_player_id)) %>%
    left_join(players, by = c("rotation_player_id" = "player_id")) %>%
    mutate(player_id = rotation_player_id) %>%
    filter(!is.na(player_name)) %>%
    group_by(player_id, player_name, rotation) %>%
    summarize(n = n()) %>%
    pivot_wider(names_from = rotation, values_from = n, values_fill = 0) %>%
    mutate(back_row = ifelse(
      sum(home_player_id5, home_player_id6)/2 > 
        0.4*sum(home_player_id2, home_player_id3, home_player_id4)/3, 
      TRUE, FALSE
    ))
}

# Convert the starting lineup into the 6 players on the court in each rotation
get_rotation_lineup <- function(lineup, system) {
  
  system <- ifelse(system == "Unknown", "5-1", system)
  
  # Define rotation patterns for each system
  rotation_map <- switch(system,
    "5-1" = list(
      Ro1 = c(Pos1 = "S1",  Pos2 = "OH1", Pos3 = "MB2", Pos4 = "OPP1", Pos5 = "OH2", Pos6 = "L"),
      Ro2 = c(Pos1 = "OH1", Pos2 = "MB2", Pos3 = "OPP1", Pos4 = "OH2", Pos5 = "L", Pos6 = "S1"),
      Ro3 = c(Pos1 = "MB2", Pos2 = "OPP1", Pos3 = "OH2", Pos4 = "MB1", Pos5 = "S1",  Pos6 = "OH1"),
      Ro4 = c(Pos1 = "OPP1", Pos2 = "OH2", Pos3 = "MB1", Pos4 = "S1",  Pos5 = "OH1", Pos6 = "L"),
      Ro5 = c(Pos1 = "OH2", Pos2 = "MB1", Pos3 = "S1",  Pos4 = "OH1", Pos5 = "L", Pos6 = "OPP1"),
      Ro6 = c(Pos1 = "MB1", Pos2 = "S1",  Pos3 = "OH1", Pos4 = "MB2", Pos5 = "OPP1", Pos6 = "OH2")
    ),
    "6-2" = list(
      Ro1 = c(Pos1 = "S1",  Pos2 = "OH1", Pos3 = "MB2", Pos4 = "OPP2", Pos5 = "OH2", Pos6 = "L"),
      Ro2 = c(Pos1 = "OH1", Pos2 = "MB2", Pos3 = "OPP2", Pos4 = "OH2", Pos5 = "L", Pos6 = "S1"),
      Ro3 = c(Pos1 = "MB2", Pos2 = "OPP2", Pos3 = "OH2", Pos4 = "MB1", Pos5 = "S1",  Pos6 = "OH1"),
      Ro4 = c(Pos1 = "S2",  Pos2 = "OH2", Pos3 = "MB1", Pos4 = "OPP1", Pos5 = "OH1", Pos6 = "L"),
      Ro5 = c(Pos1 = "OH2", Pos2 = "MB1", Pos3 = "OPP1", Pos4 = "OH1", Pos5 = "L", Pos6 = "S2"),
      Ro6 = c(Pos1 = "MB1", Pos2 = "OPP1", Pos3 = "OH1", Pos4 = "MB2", Pos5 = "S2",  Pos6 = "OH2")
    ),
    stop("Unknown system")
  )

  # Build rotation lineups
  map_dfr(names(rotation_map), function(rotation) {
    positions <- rotation_map[[rotation]]

    players <- map_chr(positions, function(pos_label) {
      lineup %>%
        filter(pos == pos_label) %>%
        pull(player_name) %>%
        first()  # In case of duplicates
    })

    tibble(
      rotation = rotation,
      Pos1 = players[1],
      Pos2 = players[2],
      Pos3 = players[3],
      Pos4 = players[4],
      Pos5 = players[5],
      Pos6 = players[6]
    )
  })
}

# Adjust the servers in this rotation lineup
adjust_servers <- function(rotation_lineup, servers, system, lineup) {
  safe_first <- function(x) if (length(x) == 0) NA_character_ else x[1]

  server_names <- servers$player_name

  server_positions <- servers %>% select(pos, player_name)
  ds1_name <- safe_first(server_positions %>% filter(pos == "DS1") %>% pull(player_name))
  ds2_name <- safe_first(server_positions %>% filter(pos == "DS2") %>% pull(player_name))
  l_name   <- safe_first(server_positions %>% filter(pos == "L") %>% pull(player_name))

  oh1_serves <- "OH1" %in% servers$pos
  oh2_serves <- "OH2" %in% servers$pos
  ds1_serves <- "DS1" %in% servers$pos
  ds2_serves <- "DS2" %in% servers$pos

  name_to_pos <- lineup %>% select(player_name, pos)

  mb1_serves <- "MB1" %in% servers$pos
  mb2_serves <- "MB2" %in% servers$pos
  both_mbs_dont_serve <- !(mb1_serves || mb2_serves)

  # Initial substitutions
  adjusted <- rotation_lineup %>%
    rowwise() %>%
    mutate(Pos1 = {
      lineup_pos <- name_to_pos %>%
        filter(player_name == Pos1) %>%
        pull(pos) %>%
        safe_first()

      if (Pos1 %in% server_names) {
        Pos1
      
      } else if (lineup_pos %in% c("MB1", "MB2")) {
        if (lineup_pos == "MB1") {
          if (!mb1_serves && !is.na(l_name) && l_name %in% server_names) {
            # MB1 doesn't serve, and L is available
            l_name
          } else {
            # Keep MB1
            Pos1
          }
        } else if (lineup_pos == "MB2") {
          if (!mb2_serves && mb1_serves && !is.na(l_name) && l_name %in% server_names) {
            # MB2 doesn't serve, MB1 does, and L is available
            l_name
          } else {
            # Keep MB2
            Pos1
          }
        } else {
          Pos1
        }

      } else if (system == "5-1") {
        case_when(
          lineup_pos == "OH1" & ds1_serves ~ ds1_name,
          lineup_pos == "OH2" & !oh1_serves & ds2_serves ~ ds2_name,
          lineup_pos == "OH2" & oh1_serves & ds1_serves ~ ds1_name,
          lineup_pos == "OPP1" & oh1_serves & oh2_serves & ds1_serves ~ ds1_name,
          lineup_pos == "OPP1" & ds2_serves ~ ds2_name,
          TRUE ~ Pos1
        )

      } else if (system == "6-2") {
        case_when(
          lineup_pos == "S1" ~ safe_first(servers %>% filter(pos == "OPP1") %>% pull(player_name)),
          lineup_pos == "S2" ~ safe_first(servers %>% filter(pos == "OPP2") %>% pull(player_name)),
          lineup_pos == "OH1" & ds1_serves ~ ds1_name,
          lineup_pos == "OH2" & !oh1_serves & ds2_serves ~ ds2_name,
          lineup_pos == "OH2" & oh1_serves & ds1_serves ~ ds1_name,
          TRUE ~ Pos1
        )

      } else {
        Pos1
      }
    }) %>%
    ungroup()

  # ------------------------
  # Final substitutions step
  # ------------------------

  # 1. Determine unused servers
  used_names <- adjusted$Pos1
  unused_servers <- servers %>%
    filter(!(player_name %in% used_names)) %>%
    pull(player_name)

  # 2. Identify Pos1s to consider for replacement (OH, OPP who don’t serve)
  name_to_pos <- name_to_pos %>% distinct()
  candidates <- adjusted %>%
    left_join(name_to_pos, by = c("Pos1" = "player_name")) %>%
    filter(pos %in% c("OH1", "OH2", "OPP1", "OPP2"), !(Pos1 %in% server_names))

  # 3. Replace non-serving OH/OPP with unused servers
  if (length(unused_servers) > 0 && nrow(candidates) > 0) {
    for (i in seq_len(nrow(candidates))) {
      old_name <- candidates$Pos1[i]
      new_name <- safe_first(setdiff(unused_servers, adjusted$Pos1))

      if (!is.na(new_name)) {
        adjusted$Pos1[adjusted$Pos1 == old_name] <- new_name
      }
    }
  }
  
  # Step 1: Identify non-server Pos1s
  non_servers <- which(!(adjusted$Pos1 %in% server_names))

  # Step 2: Get available servers not already in Pos1
  used <- adjusted$Pos1
  available <- setdiff(server_names, used)

  # Step 3: Replace at most two non-server Pos1s with available servers
  for (i in seq_along(non_servers)) {
    if (i > length(available)) break  # Stop if we run out of servers

    row_idx <- non_servers[i]
    adjusted$Pos1[row_idx] <- available[i]
  }

  return(adjusted)
}

# Adjust the back row in this rotation lineup
adjust_back_row <- function(server_adj, lineup) {
  # Create lookup for back_row by player
  lineup_backrow <- lineup %>% select(player_name, pos, back_row)

  # Define rotation order
  rotation_order <- c("Ro1", "Ro2", "Ro3", "Ro4", "Ro5", "Ro6")
  
  # Lookup table for previous rotations
  prev_rotation <- setNames(c("Ro6", "Ro1", "Ro2", "Ro3", "Ro4", "Ro5"), rotation_order)
  prev_rotation2 <- setNames(c("Ro5", "Ro6", "Ro1", "Ro2", "Ro3", "Ro4"), rotation_order)

  # Join back row info for Pos5 and Pos6
  server_adj <- server_adj %>%
    left_join(lineup_backrow, by = c("Pos6" = "player_name")) %>%
    rename(pos6 = pos, back6 = back_row) %>%
    left_join(lineup_backrow, by = c("Pos5" = "player_name")) %>%
    rename(pos5 = pos, back5 = back_row)

  # Adjust Pos6 and Pos5
  server_adj <- server_adj %>%
    rowwise() %>%
    mutate(
      Pos6 = {
        if (pos6 %in% c("L", "S1", "S2")) {
          Pos6
        } else if (pos6 %in% c("OH1", "OH2", "OPP1") && isFALSE(back6)) {
          prev_rot <- prev_rotation[[rotation]]
          sub <- server_adj %>%
            filter(rotation == prev_rot) %>%
            pull(Pos1) %>%
            first()
          if (!is.na(sub)) sub else Pos6
        } else {
          Pos6
        }
      },
      Pos5 = {
        if (pos5 %in% c("L", "S1", "S2")) {
          Pos5
        } else if (pos5 %in% c("OH1", "OH2", "OPP1") && isFALSE(back5)) {
          prev_rot2 <- prev_rotation2[[rotation]]
          sub <- server_adj %>%
            filter(rotation == prev_rot2) %>%
            pull(Pos1) %>%
            first()
          if (!is.na(sub)) sub else Pos5
        } else {
          Pos5
        }
      }
    ) %>%
    ungroup() %>%
    select(-pos6, -back6, -pos5, -back5)

  return(server_adj)
}


# Combine it all
generate_starting_lineup <- function(data, team_name) {
  system <- detect_system(data, team_name)
  lineup <- build_lineup(data, team_name, system) %>% 
    left_join(get_back_row(data, team_name) %>% select(player_id, player_name, back_row)) %>%
    mutate(back_row = ifelse(position %in% c("Outside", "Opposite"), back_row, NA))
  servers <- get_servers(data, team_name) %>%
    left_join(lineup %>% select(player_name, position, pos), by = "player_name")
  back_row <- get_back_row(data, team_name)
  rotation_lineup <- get_rotation_lineup(lineup, system)
  server_adj <- adjust_servers(rotation_lineup, servers, system, lineup)
  final_rot_lineup <- adjust_back_row(server_adj, lineup)
  
  list(
    team_short = team_name,
    system = system,
    lineup = lineup,
    servers = servers,
    back_row = back_row,
    rotation_lineup = rotation_lineup,
    server_adj = server_adj,
    final_rot_lineup = final_rot_lineup
  )
}
```

```{r}
result <- generate_starting_lineup(power, "byu")
cat("System detected:", result$system, "\n")
print(result$lineup)
print(result$servers)
print(result$final_rot_lineup)
```




WAYS TO IMPROVE THIS PROCESS
1. 

```{r}
# Use a Bradley-Terry model to quantify the advantage of playing at home
get_home_advantage <- function(data) {
  shortenings <- data %>%
    select(team, team_short) %>%
    distinct()
  
  home_away_matrix <- data %>%
    filter(!is.na(conference), !is.na(opp_conference), conference == opp_conference) %>%
    select(match_id, team, opponent, home_team) %>%
    distinct() %>%
    mutate(team_home_away = ifelse(team == home_team, "home", "away")) %>%
    left_join(shortenings, by = c("opponent" = "team")) %>%
    rename(opp_short = team_short) %>%
    left_join(shortenings, by = "team") %>%
    select(match_id, team_short, opp_short, team_home_away) %>% 
    mutate(team_home_away_value = ifelse(team_home_away == "home", 1, -1)) %>%
    pivot_wider(names_from = team_short, values_from = team_home_away_value) %>%
    group_by(match_id) %>%
    summarise_at(vars(-opp_short, -team_home_away), sum, na.rm = TRUE)
    
  home_team_win <- data %>%
    filter(!is.na(conference), !is.na(opp_conference), conference == opp_conference) %>%
    group_by(match_id) %>%
    arrange(match_id, desc(row_number())) %>%
    slice(1) %>%
    ungroup() %>%
    select(match_id, home_team, visiting_team, point_won_by) %>%
    mutate(home_team_win = ifelse(home_team == point_won_by, 1, 0)) %>%
    select(match_id, home_team_win)
  
  home_advantage <- inner_join(home_away_matrix, home_team_win, by = "match_id")
  
  bt_model <- glm(home_team_win ~ . -match_id, family = binomial, data = home_advantage)
  home_advantage <- as.numeric(coef(bt_model)["(Intercept)"])
  
  return(home_advantage)
}

# Quantify the advantage of being the sideout team
get_sideout_advantage <- function(data) {
  sideout_rate <- data %>%
    filter(skill == "Serve") %>%
    summarize(rate = mean(point_won_by != serving_team)) %>%
    pull(rate)
  
  sideout_advantage <- log(sideout_rate / (1 - sideout_rate))
  
  return(sideout_advantage)
}

# Find each team's most common starting rotation
get_starting_rotation <- function(data) {
  start_rot_data <- data %>%
    group_by(set_id) %>%
    filter(point_id == min(point_id)) %>%
    ungroup() %>%
    mutate(setter_position = ifelse(team == home_team, home_setter_position, visiting_setter_position)) %>%
    distinct(team_short, set_id, setter_position) %>%
    group_by(team_short) %>%
    summarize(
      Ro1 = sum(setter_position == 1),
      Ro2 = sum(setter_position == 2),
      Ro3 = sum(setter_position == 3),
      Ro4 = sum(setter_position == 4),
      Ro5 = sum(setter_position == 5),
      Ro6 = sum(setter_position == 6)
    ) %>%
    rowwise() %>%
    mutate(start_rot = names(across(Ro1:Ro6))[which.max(c_across(Ro1:Ro6))]) %>%
    ungroup()
  
  return(start_rot_data)
}

# Attach RAPMs to each rotation for each team and combine into one data set
combine_teams <- function(data, teamA, teamB) {
  teamA <- generate_starting_lineup(data, teamA)$final_rot_lineup %>% mutate(team = teamA) %>% select(team, everything())
  teamB <- generate_starting_lineup(data, teamB)$final_rot_lineup %>% mutate(team = teamB) %>% select(team, everything())
  
  teamA <- teamA %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos1" = "player_name")) %>% rename(rapm1 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos2" = "player_name")) %>% rename(rapm2 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos3" = "player_name")) %>% rename(rapm3 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos4" = "player_name")) %>% rename(rapm4 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos5" = "player_name")) %>% rename(rapm5 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos6" = "player_name")) %>% rename(rapm6 = RAPM) %>%
    mutate(rotation_rapm = rowSums(across(rapm1:rapm6)))
  
  teamB <- teamB %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos1" = "player_name")) %>% rename(rapm1 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos2" = "player_name")) %>% rename(rapm2 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos3" = "player_name")) %>% rename(rapm3 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos4" = "player_name")) %>% rename(rapm4 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos5" = "player_name")) %>% rename(rapm5 = RAPM) %>%
    left_join(rapm %>% select(player_name, RAPM), by = c("Pos6" = "player_name")) %>% rename(rapm6 = RAPM) %>%
    mutate(rotation_rapm = rowSums(across(rapm1:rapm6)))
  
  rotation_rapms <- rbind(teamA, teamB)
  
  return(rotation_rapms)
}
```

```{r}
get_home_advantage(power)
get_sideout_advantage(power)
get_starting_rotation(power) %>% head()
combine_teams(power, "texas", "nebraska")
```



WAYS TO IMPROVE THIS PROCESS
1. Improve computational intensity

```{r}
# SIMULATE!!
simulate <- function(data, teamA, teamB, location) {
  
  # Load starting rotations (fixed for all matches)
  start_rotationA <- get_starting_rotation(data) %>% filter(team_short == teamA) %>% pull(start_rot)
  start_rotationB <- get_starting_rotation(data) %>% filter(team_short == teamB) %>% pull(start_rot)
  
  # Get RAPM table
  rotation_rapms <- combine_teams(data, teamA, teamB)
  
  # Sideout advantage
  sideout_adv <- get_sideout_advantage(data)
  
  # Rotation order for tracking
  rotation_order <- c("Ro1", "Ro6", "Ro5", "Ro4", "Ro3", "Ro2")
  
  # Store match data
  matches <- list()
  match_num <- 1
  
  for (i in 1:1000) {
    
    # Initialize match state
    set <- 1
    scoreA <- 0
    scoreB <- 0
    setwinsA <- 0
    setwinsB <- 0
  
    # Store rally data
    rallies <- list()
    rally_num <- 1
    
    while (setwinsA < 3 && setwinsB < 3) {
      # Determine set point target
      point_target <- ifelse(set == 5, 15, 25)
    
      # Reset set scores and randomly choose serving team
      scoreA <- 0
      scoreB <- 0
      serving <- sample(c("A", "B"), size = 1)
    
      # Reset to original starting rotations
      rotationA <- start_rotationA
      rotationB <- start_rotationB
    
    # Simulate set
      while (!(max(scoreA, scoreB) >= point_target && abs(scoreA - scoreB) >= 2)) {
        # Store who is serving and rotations at start of rally
        serving_start <- serving
        rotationA_point <- rotationA
        rotationB_point <- rotationB
    
        # Get RAPMs
        rapm_A <- rotation_rapms %>% filter(team == teamA, rotation == rotationA_point) %>% pull(rotation_rapm)
        rapm_B <- rotation_rapms %>% filter(team == teamB, rotation == rotationB_point) %>% pull(rotation_rapm)
    
        # Compute probability of team A winning the point
        value <- rapm_A - rapm_B
        value <- value + ifelse(serving_start == "B", sideout_adv, -sideout_adv)
        prob_A_win <- exp(value) / (1 + exp(value))
    
        # Simulate point
        point_winner <- ifelse(runif(1) < prob_A_win, "A", "B")
    
        # Update scores
        if (point_winner == "A") {
          scoreA <- scoreA + 1
        } else {
          scoreB <- scoreB + 1
        }
    
        # Update rotations if sideout occurred
        if (point_winner != serving_start) {
          if (serving_start == "A") {
            idx <- match(rotationB, rotation_order)
            rotationB <- rotation_order[(idx %% 6) + 1]
          } else {
            idx <- match(rotationA, rotation_order)
            rotationA <- rotation_order[(idx %% 6) + 1]
          }
          serving <- point_winner
        }
    
        # Store rally
        rallies[[rally_num]] <- data.frame(
          set = set,
          rally = rally_num,
          serving_team = serving_start,
          prob_A_win = prob_A_win,
          point_winner = point_winner,
          scoreA = scoreA,
          scoreB = scoreB,
          rotationA_point = rotationA_point,
          rotationB_point = rotationB_point
        )
        rally_num <- rally_num + 1
      }
    
      # Update set wins
      if (scoreA > scoreB) {
        setwinsA <- setwinsA + 1
      } else {
        setwinsB <- setwinsB + 1
      }
    
      set <- set + 1
    }
    
    # Combine all rallies into a single data frame
    match_sim <- dplyr::bind_rows(rallies)
  
    scoreA_set1 <- match_sim %>% filter(set == 1) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreA) }
    scoreA_set2 <- match_sim %>% filter(set == 2) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreA) }
    scoreA_set3 <- match_sim %>% filter(set == 3) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreA) }
    scoreA_set4 <- match_sim %>% filter(set == 4) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreA) }
    scoreA_set5 <- match_sim %>% filter(set == 5) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreA) }
    scoreB_set1 <- match_sim %>% filter(set == 1) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreB) }
    scoreB_set2 <- match_sim %>% filter(set == 2) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreB) }
    scoreB_set3 <- match_sim %>% filter(set == 3) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreB) }
    scoreB_set4 <- match_sim %>% filter(set == 4) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreB) }
    scoreB_set5 <- match_sim %>% filter(set == 5) %>% 
      { if (nrow(.) == 0) NA else slice_tail(., n = 1) %>% pull(scoreB) }
    winner <- match_sim %>% slice_tail(n = 1) %>% pull(point_winner)
    
    # Store rally
    matches[[match_num]] <- data.frame(
      winner = winner,
      scoreA_set1 = scoreA_set1,
      scoreA_set2 = scoreA_set2,
      scoreA_set3 = scoreA_set3,
      scoreA_set4 = scoreA_set4,
      scoreA_set5 = scoreA_set5,
      scoreB_set1 = scoreB_set1,
      scoreB_set2 = scoreB_set2,
      scoreB_set3 = scoreB_set3,
      scoreB_set4 = scoreB_set4,
      scoreB_set5 = scoreB_set5
    )
    
    match_num <- match_num + 1
  }
  
  # Combine all matches into a single data frame
  probability_sim <- dplyr::bind_rows(matches) %>%
    mutate(
      winner_set1 = case_when(
        scoreA_set1 > scoreB_set1 ~ "A",
        scoreA_set1 < scoreB_set1 ~ "B",
        is.na(scoreA_set1) | is.na(scoreB_set1) ~ NA
      ),
      winner_set2 = case_when(
        scoreA_set2 > scoreB_set2 ~ "A",
        scoreA_set2 < scoreB_set2 ~ "B",
        is.na(scoreA_set2) | is.na(scoreB_set2) ~ NA
      ),
      winner_set3 = case_when(
        scoreA_set3 > scoreB_set3 ~ "A",
        scoreA_set3 < scoreB_set3 ~ "B",
        is.na(scoreA_set3) | is.na(scoreB_set3) ~ NA
      ),
      winner_set4 = case_when(
        scoreA_set4 > scoreB_set4 ~ "A",
        scoreA_set4 < scoreB_set4 ~ "B",
        is.na(scoreA_set4) | is.na(scoreB_set4) ~ NA
      ),
      winner_set5 = case_when(
        scoreA_set5 > scoreB_set5 ~ "A",
        scoreA_set5 < scoreB_set5 ~ "B",
        is.na(scoreA_set5) | is.na(scoreB_set5) ~ NA
      )
    )
  
  home_adv <- get_home_advantage(data)
  
  # if (location == "neutral") {
  #   teamA_winprob <- probability_sim %>% summarize(teamA_winprob = mean(winner == "A")) %>% pull()
  #   
  # } else if (location == "A") {
  #   neutralA_winprob <- probability_sim %>% summarize(neutralA_winprob = mean(winner == "A")) %>% pull()
  #   neutralA_logodds <- log(neutralA_winprob / (1 - neutralA_winprob))
  #   value <- neutralA_logodds + home_adv
  #   teamA_winprob <- exp(value) / (1 + exp(value))
  #   
  # } else if (location == "B") {
  #   neutralA_winprob <- probability_sim %>% summarize(neutralA_winprob = mean(winner == "A")) %>% pull()
  #   neutralA_logodds <- log(neutralA_winprob / (1 - neutralA_winprob))
  #   value <- neutralA_logodds - home_adv
  #   teamA_winprob <- exp(value) / (1 + exp(value))
  #   
  # } else {
  #   stop("Entry for location must be one of: ('A', 'B', 'neutral')")
  # }
  
  # Compute the neutral win probability first
  neutralA_winprob <- probability_sim %>% summarize(neutralA_winprob = mean(winner == "A")) %>% pull()
  neutralA_logodds <- log(neutralA_winprob / (1 - neutralA_winprob))
  
  win_probs <- tibble(
    location = c("neutral", "A", "B"),
    teamA_winprob = c(
      neutralA_winprob,
      exp(neutralA_logodds + home_adv) / (1 + exp(neutralA_logodds + home_adv)),
      exp(neutralA_logodds - home_adv) / (1 + exp(neutralA_logodds - home_adv))
    )
  )
  
  
  list(
    win_probs = win_probs,
    probability_sim = probability_sim,
    set4_percentage = mean(!is.na(probability_sim$winner_set4)),
    set5_percentage = mean(!is.na(probability_sim$winner_set5))
  )

}
```

```{r}
simulation <- simulate(power, "texas", "nebraska")
print(simulation$win_probs)
print(simulation$set4_percentage)
print(simulation$set5_percentage)
```

```{r}
# visualize the distribution of wins across sets
simulation$probability_sim %>% summarize(mean(winner == "A"))
  mutate(across(where(is.character), ~replace_na(., "none"))) %>%
  summarize(across(starts_with("winner_set"), ~sum(. == "A"), .names = "A{.col}"),
            across(starts_with("winner_set"), ~sum(. == "B"), .names = "B{.col}")) %>%
  pivot_longer(everything(), names_to = "key", values_to = "count") %>%
  mutate(
    team = substr(key, 1, 1),
    set_num = substr(key, nchar(key), nchar(key)),
    set = factor(set_num, levels = as.character(1:5), labels = paste("Set", 1:5)),
    signed_count = ifelse(team == "A", count, -count),
    abs_signed = abs(signed_count)
  ) %>%
  group_by(set_num) %>%
  mutate(p_win_set = abs_signed / sum(abs_signed)) %>%
  ungroup() %>%
  ggplot(aes(x = set, y = signed_count, fill = team)) +
  geom_col(width = 0.7, position = "identity") +
  geom_text(aes(y = signed_count / 2, label = scales::percent(p_win_set, accuracy = 1)), color = "white") +
  scale_fill_manual(values = c("A" = "#CC5500", "B" = "#E41C38")) +
  geom_hline(yintercept = 0, color = "black") +
  labs(title = "Set Wins by Team", x = NULL, y = "Number of Set Wins", fill = "Team") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


